Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
merge

\begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & (integer | null) & 60570 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
\AT{More on these functions?}

@@

 \begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Arguments & \% Total & Cumulative \%\\
\midrule
 character[] & character & 183 & 2.05 & 2.05\\
 \textasciicircum{}logical & null & 158 & 1.77 & 3.81\\
 character & null & 117 & 1.31 & 5.12\\
 double & null & 107 & 1.20 & 6.32\\
 double[] & double & 100 & 1.12 & 7.44\\
 class<data.frame> & ${\bf class} \langle data.frame, tbl, tbl\_df \rangle$ & 63 & 0.70 & 8.14\\
 \bottomrule
\end{tabular}

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Most common failing contract checks. Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
 \end{table}

%
%
\subsubsection{Patterns in Failed Contract Assertions}

While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.

We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.

>>>>

\begin{tabular}{llrrr}
\toprule
Actual Type & Expected Type & \# Failed Assertions & \% Total & Cumulative \%\\
\midrule
double[] & class<`matrix`> & 371896 & 27.37 & 27.37\\
double[] & integer[] & 117212 & 8.62 & 35.99\\
character & class<`bignum`> | raw[] & 100100 & 7.37 & 43.36\\
double & integer | null & 60570 & 4.46 & 47.81\\
double[] & class<`timeSeries`> & 58872 & 4.33 & 52.15\\
\addlinespace
class<matrix> & class<`timeSeries`> & 55005 & 4.05 & 56.19\\
double[] & double & 53338 & 3.92 & 60.12\\
double & class<`data.frame`> & 32553 & 2.40 & 62.51\\
double[] & class<`data.frame`> & 31576 & 2.32 & 64.84\\
double & integer & 20586 & 1.51 & 66.35\\
\bottomrule
\end{tabular}
% END Autogenerated
\label{tbl:fail-breakdown-0}
\caption{Top types of contract assertion failures.}
\end{table}

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
These are functions which were under-tested, as calls to these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the core corpus to infer types.

@@

% END Autogenerated
 \label{tbl:fail-breakdown-1}
 \caption{Most common failing contract checks by type. Note that for this table, we collapsed multiple identical failed assertions into one, meaning that single arguments with multiple (identical) failed assertions are only counted once.}
 \end{table}

%
%
% \subsubsection{Patterns in Failed Contract Assertions}

% While Table~\ref{tbl:fail-breakdown-1} revealed no glaring patterns in failed assertions, there are a few subtle issues that we will discuss.
% First, \PERCFAILEDVECTOSCA of failed checks broken down by parameter position failed due to vectors being passed where scalars were expected.

% We observed an interesting corner case and quirk of R in looking at functions with values that default to NA.
% In R, recall that there is a different type of NA for each primitive, and when the literal value \code{NA} appears in code, R assumes it to be the logical NA.
% For the purposes of our analysis, this has the effect of introducing $\string^{\bf lgl}$ types where they (probably) don't belong, particularly when programmers fail to effectively test their code.
% For example, a function header like \code{function(x, y=NA)} will have $\string^{\bf lgl}$ as part of the type of parameter \code{y}.
% We tried to control for this in our analysis, and indeed $\string^{\bf lgl}$ is a subtype of all NA-able numeric types.
